---
title: "FP_Analysis"
author: "Jakob Balkovec"
date: "`r Sys.Date()`"
output: html_document
---

# 0/1 Knapsack Problem
## Data Analysis and Graph Generation

### Reading the Data

We read in the data from the Excel file, manipulate it slighty so it fits our requirements. We created tuples with 5 variables each, representing a single run, 150 runs total.

```{r, echo=TRUE, results='hide'}
# Read in the data
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(knitr))
suppressPackageStartupMessages(library(kableExtra))
suppressPackageStartupMessages(library(pander))

file_name <- "excel_data/Results.xlsx"
sheet_name <- "Raw"

data <- read_excel(file_name, sheet = sheet_name)

# Tuples format:
#
# {(Run #, Time, Capacity, Solution, Size, Type)}
#
# Size: 150 Tuples, 6 variables each
#
tuples <- data %>%
  select(Run, DP, CapacityDP, SolutionDP, Size) %>%
  rename(Time = DP, Capacity = CapacityDP, Solution = SolutionDP) %>%
  mutate(Type = "DP") %>%
  bind_rows(
    data %>%
      select(Run, Memo, CapacityMemo, SolutionMemo, Size) %>%
      rename(Time = Memo, Capacity = CapacityMemo, Solution = SolutionMemo) %>%
      mutate(Type = "Memo"),
    data %>%
      select(Run, Recursive, CapacityRecursive, SolutionRecursive, Size) %>%
      rename(Time = Recursive, Capacity = CapacityRecursive, Solution = SolutionRecursive) %>%
      mutate(Type = "Recursive")
  )

tuples_clean_rec <- tuples %>% 
  filter(Type == "Recursive", !is.na(Capacity), !is.infinite(Capacity), Capacity > 0)

tuples_clean_DP <- tuples %>% 
  filter(Type == "DP", !is.na(Capacity), !is.infinite(Capacity), Capacity > 0)

tuples_clean_memo <- tuples %>% 
  filter(Type == "Memo", !is.na(Capacity), !is.infinite(Capacity), Capacity > 0)

combined_tuples <- bind_rows(tuples_clean_rec, tuples_clean_DP, tuples_clean_memo)
```

## Exploratory Analysis on the Dataset

### Summary Statistics
```{r, echo=TRUE, results='show'}

custom_summary <- function(df) {
  df %>%
    summarise(
      Min = min(Capacity, na.rm = TRUE),
      Q1 = quantile(Capacity, 0.25, na.rm = TRUE),
      Median = median(Capacity, na.rm = TRUE),
      Mean = mean(Capacity, na.rm = TRUE),
      Q3 = quantile(Capacity, 0.75, na.rm = TRUE),
      Max = max(Capacity, na.rm = TRUE)
    )
}

table <- function(tuples, name) {
  kable(custom_summary(tuples), "html", caption = paste("Custom Summary of", name, "Tuples")) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(1:6, width = "100px")
}
summary_rec <- as.data.frame(summary(tuples_clean_rec))
summary_dp <- as.data.frame(summary(tuples_clean_DP))
summary_memo <- as.data.frame(summary(tuples_clean_memo))

table(tuples_clean_rec, "Recursive")
table(tuples_clean_DP, "DP")
table(tuples_clean_memo, "Memo")
```
###  Distribution Analysis
```{r, echo=TRUE, results='show'}
tuples_clean_rec$Type <- "Recursive"
tuples_clean_DP$Type <- "DP"
tuples_clean_memo$Type <- "Memo"

suppressWarnings(
ggplot(combined_tuples, aes(x = Capacity, fill = Type)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity", aes(y = after_stat(density))) +  
  geom_density(alpha = 0.3, color = "black", adjust = 1) +
  facet_wrap(~ Type, scales = "free_y") + 
  theme_minimal() +
  labs(title = "Capacity Distribution by Algorithm Type", x = "Capacity", y = "Density") +
  scale_x_continuous(limits = c(-500, 500), breaks = seq(-500, 500, by = 100)) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 6),  # Adjust the size of the x-axis ticks
    axis.text.y = element_text(size = 6)   # Adjust the size of the y-axis ticks
  )
)
```

```{r, echo=TRUE, results='show'}
# Boxplot for solution comparison
ggplot(combined_tuples, aes(x = Type, y = Solution, fill = Type)) +
  geom_boxplot() +  # Create the boxplot
  theme_minimal() +
  labs(title = "Solution Comparison by Algorithm Type", x = "Algorithm Type", y = "Solution") +
  theme(legend.position = "none")
```

### Correlation Analysis
```{r, echo=TRUE, results='show'}
# Calculate correlations
tuples %>%
  group_by(Type) %>%
  summarize(correlation = cor(Capacity, Solution, use = "complete.obs"))
```

### Comparing Algorithm Performance
```{r, echo=TRUE, results='show'}
# Calculate the minimum capacity
min_capacity <- min(combined_tuples$Capacity)

ggplot(combined_tuples, aes(x = Type, y = Solution, fill = Type)) +
  stat_summary(fun = mean, geom = "bar", color = "black") + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
  geom_jitter(aes(color = Type), width = 0.2, alpha = 0.5) +  # Map color to Type
  scale_color_manual(values = c("DP" = "tomato", "Memo" = "springgreen3", "Recursive" = "royalblue2")) +  # Set specific colors
  theme_minimal() +
  labs(title = "Solution by Algorithm Type with Data Points", 
       x = "Algorithm Type", y = "Solution") +
  theme(plot.title = element_text(hjust = 0.5, size = 16), 
        axis.text = element_text(size = 12),  
        axis.title = element_text(size = 14)) +
  scale_y_continuous(limits = c(min_capacity, NA))
```

```{r}
ggplot(combined_tuples, aes(x = Type, y = Solution, fill = Type)) +
  stat_summary(fun = mean, geom = "bar", color = "black") + 
  stat_summary(fun.data = mean_cl_normal, geom = "errorbar", width = 0.2, color = "black") +
  geom_jitter(color = "black", width = 0.2, alpha = 0.5) +
  theme_minimal() +
  labs(title = "Average Solution by Algorithm Type with Data Points", 
       x = "Algorithm Type", y = "Average Solution") +
  theme(plot.title = element_text(hjust = 0.5, size = 16), 
        axis.text = element_text(size = 12),  
        axis.title = element_text(size = 14))

# Bins show mean, points are individual solutions, error bars are error bars
```

### Performance by Problem Size
```{r, echo=TRUE, results='show'}
# From XS to XL, recursive fix
ggplot(combined_tuples, aes(x = Size, y = Capacity, fill = Type)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Capacity by Problem Size and Algorithm Type", x = "Problem Size", y = "Capacity")
```
```{r}
ggplot(combined_tuples, aes(x = Size, y = Solution, fill = Type)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Solution by Problem Size and Algorithm Type", x = "Problem Size", y = "Capacity")
```

### Outlier Detection
```{r, echo=TRUE, results='show'}
# Outliars in Execution time, better graph
ggplot(combined_tuples, aes(x = Type, y = Time, fill=Type)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Outliers in Time by Algorithm Type", x = "Algorithm Type", y = "Time") +
  theme(plot.title = element_text(hjust = 0.5)) +
  facet_wrap(~ Type, scales = "free_y")
```

### Relationship Between Runtime and Solution Quality
```{r, echo=TRUE, results='show'}
# Plot for DP
ggplot(tuples %>% filter(Type == "DP"), aes(x = Time, y = Solution)) +
  geom_point(color = "tomato") +
  theme_minimal() +
  labs(title = "DP Runtime vs Solution Quality", x = "DP Runtime", y = "Solution Quality")
```

```{r, echo=TRUE, results='show'}
# Plot for Memo
ggplot(tuples %>% filter(Type == "Memo"), aes(x = Time, y = Solution)) +
  geom_point(color = "springgreen3") +
  theme_minimal() +
  labs(title = "Memo Runtime vs Solution Quality", x = "Memo Runtime", y = "Solution Quality")
```

```{r, echo=TRUE, results='show'}
# Plot for Recursive
ggplot(tuples %>% filter(Type == "Recursive"), aes(x = Time, y = Solution)) +
  geom_point(color = "royalblue2") +
  theme_minimal() +
  labs(title = "Recursive Runtime vs Solution Quality", x = "Recursive Runtime", y = "Solution Quality")
```
### Plot with Subplots
```{r, echo=TRUE, results='show'}
ggplot(tuples, aes(x = Time, y = Solution)) +
  geom_point(aes(color = Type)) +  # Color by Type
  theme_minimal() +
  labs(title = "Runtime vs Solution Quality by Algorithm", 
       x = "Runtime", 
       y = "Solution Quality") +
  facet_wrap(~ Type, scales = "free") +  # Create separate plots for each Type
  scale_color_manual(values = c("DP" = "tomato", "Memo" = "springgreen3", "Recursive" = "royalblue2")) +
  theme(plot.title = element_text(hjust = 0.5, size = 16), 
        axis.text = element_text(size = 12),  
        axis.title = element_text(size = 14))
```

### Relationship Between Runtime and Solution Quality
```{r, echo=TRUE, results='show'}
# Plot for DP
ggplot(tuples %>% filter(Type == "DP"), aes(x = Time, y = Capacity)) +
  geom_point(aes(color = Size)) +
  theme_minimal() +
  labs(title = "DP Runtime vs Capacity", x = "DP Runtime", y = "Capacity")
```

```{r, echo=TRUE, results='show'}
# Plot for Memo
ggplot(tuples %>% filter(Type == "Memo"), aes(x = Time, y = Capacity)) +
  geom_point(aes(color = Size)) +
  theme_minimal() +
  labs(title = "Memo Runtime vs Capacity", x = "Memo Runtime", y = "Capacity")
```

```{r, echo=TRUE, results='show'}
# Plot for Recursive
print(tuples_clean_rec)
ggplot(tuples_clean_rec, aes(x = Time, y = Capacity)) +
  geom_point(aes(color = Size)) +
  theme_minimal() +
  labs(title = "Recursive Runtime vs Capacity", x = "Recursive Runtime", y = "Capacity")
```

### Relationships
```{r, echo=TRUE, results='show'}

# ----------- DP -----------
# Relationship: Quadratic model
#
# Quadratic relationship between DP Execution Time and Capacity
# EQN: a * (Capacity^2) + b * (Capacity) + c

lm_dp_quad <- lm(Time ~ Capacity + I(Capacity^2), data = tuples_clean_DP)

intercept_dp_quad <- coef(lm_dp_quad)[1]
slope_dp_quad <- coef(lm_dp_quad)[2]
quad_slope_dp <- coef(lm_dp_quad)[3]
r_squared_dp_quad <- summary(lm_dp_quad)$r.squared

reg_eq_dp_quad <- paste("y = ", round(quad_slope_dp, 4), " Capacity² + ", round(slope_dp_quad, 2), " * Capacity + ", round(intercept_dp_quad, 2),
                        "\n", "R² = ", round(r_squared_dp_quad, 3))

ggplot(tuples %>% filter(Type == "DP"), aes(x = Capacity, y = Time)) +
  geom_point(color = "tomato") +  # Plot the data points
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "brown") +  # Quadratic fit
  theme_minimal() +
  labs(title = "DP Runtime vs Capacity with Quadratic Regression", 
       x = "Capacity", y = "DP Runtime") +
  theme(plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = max(tuples_clean_DP$Capacity) * 0.5, 
           y = max(tuples_clean_DP$Time) * 0.9,
           label = reg_eq_dp_quad, color = "black", size = 5, hjust = 0)
# ----------- DP -----------
```

```{r, echo=TRUE, results='show'}
# ---------- Memo ----------
# Relationship: Quadratic model
#
# Quadratic relationship between Memo Execution Time and Capacity
# EQN: a * (Capacity^2) + b * (Capacity) + c

lm_memo_quad <- lm(Time ~ Capacity + I(Capacity^2), data = tuples_clean_memo)

intercept_memo_quad <- coef(lm_memo_quad)[1]
slope_memo_quad <- coef(lm_memo_quad)[2]
quad_slope_memo <- coef(lm_memo_quad)[3]
r_squared_memo_quad <- summary(lm_memo_quad)$r.squared

reg_eq_memo_quad <- paste("y = ", round(quad_slope_memo, 4), " * Capacity² + ", round(slope_memo_quad, 2), " * Capacity + ", round(intercept_memo_quad, 2),
                          "\n", "R² = ", 1 - round(r_squared_memo_quad, 3))

print(reg_eq_memo_quad)

ggplot(tuples %>% filter(Type == "Memo"), aes(x = Capacity, y = Time)) +
  geom_point(color = "springgreen3") +  # Plot the data points
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), se = FALSE, color = "brown") + 
  theme_minimal() +
  labs(title = "Memo Runtime vs Capacity with Quadratic Regression", 
       x = "Capacity", y = "Memo Runtime") +
  theme(plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = max(tuples_clean_memo$Capacity) * 0.5, 
           y = max(tuples_clean_memo$Time) * 0.9,
           label = reg_eq_memo_quad, color = "black", size = 5, hjust = 0)

# ---------- Memo ----------
```

```{r, echo=TRUE, results='show'}
# -------- Recursive -------
# Relationship:
#
# Log relationship between Recursive Execution Time and the Capacity
# EQN: a * log(Capacity) + b
#
log_model <- lm(Time ~ log(Capacity), data = tuples_clean_rec)

intercept_rec <- coef(log_model)[1]
slope_rec <- coef(log_model)[2]
r_squared_rec <- summary(log_model)$r.squared

reg_eq_rec <- paste("y = ", round(slope_rec, 2), " * log(Capacity)  - ", round(intercept_rec * (-1), 2), 
                "\n", "R² = ", 1 - round(r_squared_rec, 3))

ggplot(tuples_clean_rec, aes(x = Time, y = Capacity)) +
  geom_point(color = "royalblue2") +  # Data points
  geom_smooth(method = "lm", formula = y ~ log(x), se = FALSE, color = "brown", size = 0.5) +
  theme_minimal() +
  labs(title = "Recursive Runtime vs Capacity with Logarithmic Regression",
       x = "Recursive Runtime", y = "Capacity") +
  theme(plot.title = element_text(hjust = 0.5)) +
  annotate("text", x = max(tuples_clean_rec$Time) * 0.5, y = max(tuples_clean_rec$Capacity) * 0.9,
           label = reg_eq_rec, color = "black", size = 5, hjust = 0, vjust = 1)
# -------- Recursive -------
```

### Summary Table
```{r, echo=TRUE, results='show'}
# Prettify

tuples %>%
  group_by(Type) %>%
  summarise(
    Mean_Capacity = mean(Capacity),
    Median_Capacity = median(Capacity),
    SD_Capacity = sd(Capacity),
    Mean_Time = mean(Time),
    Median_Time = median(Time),
    SD_Time = sd(Time)
  ) %>%
  kable("html", caption = "Summary Statistics by Algorithm Type") %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover", "condensed", "responsive")) %>%
  column_spec(2:7, width = "100px")
```

***
Jakob Balkovec, Seattle University, 2024

